Modeling in Calculus: Early and Often
======================================
Materials from a 2012 MMA PREP Workshop
------------------

```{r, echo=FALSE}
displayFile <- function(path, text=path, type=c('html','Rmd')) {
  result <- paste(text, paste("[[", type, "] ](", path, ".", type, ")", sep="", collapse="" ) )
  return(result)
  }
```

This is where we will post materials from our 2012 MAA PREP Workshop.  Stay tuned.

### Monday: *Introductions and Overview*

[Calvin College Campus Map](http://www.calvin.edu/map/)

[Concept Inventories Kick-Off](http://mosaic-web.org/web/KickOff/Presentations/concept-inventory.pdf)
[Calculus-Related Modeling](http://rpubs.com/dtkaplan/792)
[Test Anxiety and Course Performance](https://dl.dropbox.com/u/5098197/PREP2012/Concept-Inventory-Statistics-2012-07.pdf)

### Tuesday: *Calculus in 1 Variable*

#### Mathematics and Modeling
[Our Definitions of Modeling](http://rpubs.com/dtkaplan/818)
[Introduction to Modeling](http://rpubs.com/dtkaplan/808)

 * `r I(displayFile("Topics/Tides/tides-hawaii", "Tides in Hawaii"))`
 * `r I(displayFile("Topics/Optimization/Dog-Calculus", "Dog Calculus is Economics"))`

### Wednesday: *Calculus in Multiple Variables* 

### Thursday: *Dynamics and Differential Equations*


<!-- ### Files by topic -->
```{r, results='asis', echo=FALSE, eval=FALSE}
directories <- dir('Topics')
files <- c()
for (d in directories){
  cat(paste('\n####', d, "\n"))
  files <- dir(paste('Topics/',d, sep=""))
  for (f in files) {
    if (grepl('.html', f) ) {
      cat(paste(' * [',d, "/", f,']', sep=""))
      cat(paste('(',d, "/", f,')\n', sep=""))
      }
  }
}
```