Modeling Definitions
====================

Definitions of modeling provided **before** the workshop by the PREP 2012 participants.

* Modeling is the distillation of a complex phenomenon into **fundamental** and **quantifiable** pieces that allow us to make predictions.

* Modeling is the best attempt to explain a process.

* Representing or characterizing reality for some predictive purpose.

* Modeling is physical phenomena (including chemistry, biology, etc.) expressed and analyzed through mathematics.

* Using mathematics to combine what we have observed with what we are interested in finding out to give us information about how the world works.

* Collecting, interpreting, and presenting data in a visual, concrete form using patterns (equations) that repeat in many instances.

* Applying mathematics to solve an interesting or practical problem.

* Identifying (or guessing) key factors in a real-world situation and coming up with a systematic way to make predictions based on those.

* Using mathematics to describe phenomena (a situation) and to make predictions.

* Using real-life problems to teach/demonstrate/conceptualize topics and ideas in calculus and mathematics.  Bringing the real-world into the classroom. 

* Modeling is the process of finding a good mathematical explanation or representation for some observed phenomenon.

* Modeling is the process of constructing, testing, and applying mathematical descriptions to physical and conceptual systems, with the intent of description and/or prediction.

* All models are wrong, but some are useful.  How do we teach students to think in terms of models?

* Modeling is using mathematics to match the past and predict the future.

* Modeling is a mathematical framework for predicting outcomes of real-world problems.

* Modeling is using mathematical theory to interpret phenomena of science and engineering.

* Modeling is converting a real-world problem into a math problem.

* Modeling is the process of creating a mathematical construct to solve a real-world problem.

* Modeling is representing reality in a simple enough way to learn something useful.

### A Modeling Definition WordCloud

```{r label="wordcloud",message=FALSE,echo=FALSE,fig.out="6in"}
set.seed(13)
require(wordcloud)
require(tm)
us <- URISource("/Users/kaplan/kaplanfiles/github/PREP-2012/Participants/modeling-words-pre.txt")
foo = Corpus(us)
foo2 = tm_map(foo, removePunctuation)
foo2 = tm_map(foo2, function(x)removeWords(x,stopwords()))
tdm = TermDocumentMatrix(foo2)
m <- as.matrix(tdm)
v <- sort(rowSums(m),decreasing=TRUE)
d = data.frame(word=names(v),freq=v)
wordcloud(d$word, d$freq, min.freq=1)
```
### At the end of the workshop

```{r label="wordcloud2",message=FALSE,echo=FALSE,fig.out="6in"}
set.seed(128)
us <- URISource("/Users/kaplan/kaplanfiles/github/PREP-2012/Participants/modeling-words-post.txt")
foo = Corpus(us)
foo2 = tm_map(foo, removePunctuation)
foo2 = tm_map(foo2, function(x)removeWords(x,stopwords()))
tdm = TermDocumentMatrix(foo2)
m <- as.matrix(tdm)
v <- sort(rowSums(m),decreasing=TRUE)
d = data.frame(word=names(v),freq=v)
wordcloud(d$word, d$freq, min.freq=1)
```


### Calculus Definitions

* Calculus = approximation and limits to get accurate answers.

* Modeling using small amounts of change.

* Calculus -- the study of change and the summation of infinite quantities and their application.

* Calculus is the study of how fast changing quantities change and/or accumulate.

* Calculus is the study of change.

* Calculus: study of change.
